Helixer: Cross-species Gene Annotation Of Large Eucaryotic Genomes Using Deep Learning

### 2020-05-29 paper/main.tex
minor changes

\subsection{Metrics}
We mainly use two metrics to judge model performance. The {\it Genic F1} is our 
primary metric that provides the most comprehensive picture of annotation quality 
in one number. The {\it Subgenic F1} is similar to the Genic F1, except that it does 
not take UTR predictions into account. This is done for comparability with AUGUSTUS 
as further explained in Section \ref{sec:augustus}. Both metrics work by summing up 
True Positives (TP), False Positives (FP) and False Negatives (FN) of each 
considered class before calculating precision and recall from these values and 
combining the two into a F1 score. Neither metric takes the values of the 
intergenic class into account, as this class is very abundant and appears to be by 
far the easiest to predict. (See Table \ref{tab:statistics_all} for the class 
distribution and Tables S\ref{suptab:generalization_animals} 
and S\ref{suptab:generalization_plants} for a more in-depth report our model performance 
including the intergenic class). In the case of the Genic F1, this means that 
effectively only the TP of the intergenic class are disregarded. The metrics 
essentially provide a weighted mean of the performances in the considered classes,
with weights proportional to class frequency.

@@

Orthologs of highly conserved generally single-copy genes were identified in each genome using
BUSCO. The {\it viridiplantae} set was used for plants and the {\it metazoa} set for animals.
The training genbank files were generated directly by BUSCO, by utilizing the ``--long"  % inline code?
parameter. For plants, the entire retraining could be performed as above; however, for animals
we randomly selected only half the BUSCO-generated training set, which resulted in a training set 
size and runtime comparable to the plants (~400 genes and several days per species). For animals
the training was carried out with the subsetted training file and using the following scripts provided
by AUGUSTUS. An untrained model was setup with ``new\_species.pl", and the model was fit by running
"etraining" before and after the major hyper-parameter optimization with ``optimize\_augustus.pl".
Using the trained model for each species, we ran the main prediction ("augustus") with ``--UTR=off"
and ``--gff3=on". The gffs produced by AUGUSTUS were imported into GeenuFF and exported as h5
files in the same manor as the reference, allowing direct comparison with reference and 
Helixer predictions. While tenable for some 237 species, this method allowed neither training
nor prediction of UTR regions with AUGUSTUS, so the metric Subgenic F1, which excludes UTR,
was used for all comparisons between AUGUSTUS and Helixer.

\subsection{Evaluation against independent RNAseq data}
\label{sec:rnaseq}
The reference annotations were created with existing tools and largely with a pipeline incorporating
{\it de novo} gene predictions with RNAseq and homology data. The references, like any
data, are expected to contain errors so we chose to use RNAseq data for independent evaluation.
In plants there is the additional concern that the references may share biases with
AUGUSTUS or HMMs as AUGUSTUS was used as the {\it de novo} gene caller for the reference of
many plant species % (e.g. {\it Arabidopsis lyrata}, \cite{rawat2015improving}; 
%{\it Asparagus officinalis}, 
%\cite{harkess2017asparagus}; and {\it Chenopodium quinoa}, \cite{jarvis2017genome})  
and other HMM-based tools for many more. %(e.g. {\it Citrus clementina}, \cite{wu2018genomics}; and
%{\it Capsella grandiflora}, \cite{slotte2013capsella}). 
The Ensemble animal dataset by and large
used the Ensembl annotation pipeline, which includes no HMM-based tool.

@@

% todo, add versions to everything below
For each of these six species the following search was performed on Sequence Read Archive
`(("<species name>"[Organism] OR <species name>[All Fields]) AND stranded[All Fields]) 
AND ("biomol rna"[Properties] AND "library layout paired"[Properties])'. In so far as 
more than 50 samples were identified, every Nth sample was selected so that under 50
samples were chosen for further processing (Table S\ref{suptab:rnaseq_samples}).
Each sample was quality controlled with FastQC (www.bioinformatics.babraham.ac.uk/projects/fastqc/), before and after trimming
with Trimmomatic \citep{bolger2014trimmomatic}, and then mapped to the genome with Hisat2 \citep{kim2019graph}. 
Samtools \citep{li2009sequence} was used for the sorting and indexing that was necessary both for
RNAseq mapped metric collection with PicardTools (http://broadinstitute.github.io/picard/) and custom downstream 
processing below. MultiQC \citep{ewels2016multiqc} was used to summarize the results. The above was 
automated and the results visualized with the code here (... make available and link to RNAseq stuff).
Samples were filtered to those that had {\it relatively} high mapping rates, high mapping to
exonic regions relative to other regions; that had low 3' bias, normal FastQC and 
Trimmomatic stats, and that
were confirmed to be stranded with 2nd strand as sense strand. If more remained, seven
of the high quality samples were selected randomly.

Finally, the selected, quality, mapped RNAseq samples were merged and quantified to
get the coverage (number of reads matching; i.e. cigar =, M or X),
and spliced coverage (number of reads with gap or splice; i.e. cigar N or D)
for every base pair in the genomes as implemented in the "training\_rnaseq.py"
script.

>>>>

\subsection{Metrics}
\label{sec:metrics}
We mainly use two metrics to judge model performance. The {\it Genic F1} is our 
primary metric that provides the most comprehensive picture of annotation quality 
in one number. The {\it Subgenic F1} is similar to the Genic F1, except that it does 
not take UTR predictions into account. This is done for comparability with AUGUSTUS 
as further explained in Section \ref{sec:augustus}. Both metrics work by summing up 
True Positives (TP), False Positives (FP) and False Negatives (FN) of each 
considered class before calculating precision and recall from these values and 
combining the two into a F1 score. Neither metric takes the values of the 
intergenic class into account, as this class is very abundant and appears to be by 
far the easiest to predict. (See Table \ref{tab:statistics_all} for the class 
distribution and Tables S\ref{suptab:generalization_animals} 
and S\ref{suptab:generalization_plants} for a more in-depth report our model performance 
including the intergenic class). In the case of the Genic F1, this means that 
effectively only the TP of the intergenic class are disregarded. The metrics 
essentially provide a weighted mean of the performances in the considered classes,
with weights proportional to class frequency.

@@

Orthologs of highly conserved generally single-copy genes were identified in each genome using
BUSCO. The {\it viridiplantae} set was used for plants and the {\it metazoa} set for animals.
The training genbank files were generated directly by BUSCO, by utilizing the ``--long"  % inline code?
parameter. For plants, the entire retraining could be performed as above; however, for animals
we randomly selected only half the BUSCO-generated training set, which resulted in a training set 
size and runtime comparable to the plants (about 400 genes and several days per species). For animals
the training was carried out with the subsetted training file and using the following scripts provided
by AUGUSTUS. An untrained model was setup with ``new\_species.pl", and the model was fit by running
"etraining" before and after the major hyper-parameter optimization with ``optimize\_augustus.pl".
Using the trained model for each species, we ran the main prediction ("augustus") with ``--UTR=off"
and ``--gff3=on". The gffs produced by AUGUSTUS were imported into GeenuFF and exported as HDF5
files in the same manor as the reference, allowing for the direct comparison with both the reference and 
Helixer predictions. While tenable for some 237 species, this method allowed neither training
nor prediction of UTR regions with AUGUSTUS, so the metric Subgenic F1, which disregards the class UTR,
was used for all comparisons between AUGUSTUS and Helixer (see Section \ref{sec:metrics} for more details) .

\subsection{Evaluation against independent RNAseq data}
\label{sec:rnaseq}
The reference annotations were created with existing tools and largely with a pipeline incorporating
{\it de novo} gene predictions with RNAseq and homology data. As the references, like any
data, are expected to contain errors we chose to use RNAseq data for an independent evaluation.
In plants there is the additional concern that the references may share biases with
AUGUSTUS or HMMs as AUGUSTUS was used as the {\it de novo} gene caller for the reference of
many plant species % (e.g. {\it Arabidopsis lyrata}, \cite{rawat2015improving}; 
%{\it Asparagus officinalis}, 
%\cite{harkess2017asparagus}; and {\it Chenopodium quinoa}, \cite{jarvis2017genome})  
and other HMM-based tools for many more. %(e.g. {\it Citrus clementina}, \cite{wu2018genomics}; and
%{\it Capsella grandiflora}, \cite{slotte2013capsella}). 
The Ensemble animal dataset by and large
used the Ensembl annotation pipeline, which includes no HMM-based tool.

@@

% todo, add versions to everything below
For each of these six species the following search was performed on Sequence Read Archive
`(("<species name>"[Organism] OR <species name>[All Fields]) AND stranded[All Fields]) 
AND ("biomol rna"[Properties] AND "library layout paired"[Properties])'. In so far as 
more than 50 samples were identified and every Nth sample was selected so that under 50
samples were chosen for further processing (see Table S\ref{suptab:rnaseq_samples}).
Each sample was quality controlled with FastQC (www.bioinformatics.babraham.ac.uk/projects/fastqc/), before and after trimming
with Trimmomatic \citep{bolger2014trimmomatic}, and then mapped to the genome with Hisat2 \citep{kim2019graph}. 
Samtools \citep{li2009sequence} was used for the sorting and indexing that was necessary both for
RNAseq mapped metric collection with PicardTools (http://broadinstitute.github.io/picard/) and custom downstream 
processing below. MultiQC \citep{ewels2016multiqc} was used to summarize the results. The above was 
automated and the results visualized with the code here (... make available and link to RNAseq stuff).
The samples were filtered based on many different metrics. Relatively high mapping rates, high mapping to
exonic regions relative to other regions, low 3' bias, normal FastQC, Trimmomatic stats and a confirmation
to be stranded with 2nd strand as sense strand were all taken into account. If more remained, seven
of the high quality samples were selected randomly.

Finally, the selected and mapped RNAseq samples were merged and quantified to
get the coverage (number of reads matching; i.e. cigar =, M or X),
and spliced coverage (number of reads with gap or splice; i.e. cigar N or D)
for every base pair in the genomes as implemented in the "training\_rnaseq.py"
script.
